---
title: "Veg_Analysis"
format: html
date created: 2/22/2024
editor: visual
---

#Vegetation Analysis - Izzy Funes

Load the Primary Dataframe

```{r}
load("CRAM_Plant_Primary.Rdata")
```

Load packages

```{r}
library(tidyverse)
library(readxl)
library(vegan)
```

```{r}
Veg_Primary <- CRAM_PlantJoin 

Species_List <- Veg_Primary |>
  select(Final_ID, ecramid)


Species_wide <- Species_List |> 
  mutate(Present = 1) |> 
  pivot_wider(names_from = Final_ID, values_from = Present, values_fn = sum, values_fill = 0)
     #takes columns and turns them into rows which shortens the number of rows (makes into dataframe format)
Species_wide


 Species_matrix <- Species_wide[2:ncol(Species_wide)] |> 
  as.matrix()

rownames(Species_matrix) <- Species_wide$ecramid


Species_matrix
#Species_Transpose <- t(Species_List) 
#View(Species_matrix)
```

```{r}
Species_List |>
  group_by(ecramid) |>
  count()
```

```{r}
#ggplot(Species_List) + 
  #aes(x = ecramid, y = , fill = Final_ID) +
  #geom_bar() + 
  #labs(x = "ECRAM site IDs", 
       #y = "Species Abundance") +
  #facet_grid(ecramid~.)
  
```

```{r}
#Try Jaccard or Bray-Curtis index

transpose <- t(Species_matrix)
1 - vegdist(transpose, method = "jaccard")

1 - vegdist(transpose, method = "bray")
```

Do an NMDS, PCA or PCoA

Step 1

```{r}
ecramid.nmds <- metaMDS(Species_matrix, distance = "man")
# same options for distance as vegdist


```

plot for preliminary assessment

```{r}
plot(ecramid.nmds, type = "t", display = "species")

plot(ecramid.nmds, type = "p", display = c("site", "species"))
```

Extract scores

Insert fire history data

```{r}
Fire_History_load <- read_xlsx("NEW_CRAM_SITES4.xlsx")

Fire_History_Clean <- Fire_History_load |>
  select(ecramid,`Fire/Control`, `Pre/Post_Fire`, FIRE_NAME)

Fire_Join <- left_join(Species_List, Fire_History_Clean)

```

native/non-native is "invasive" score (0/1)

filter in site data (comparing pre-fire and post-fire)

```{r}
scores(ecramid.nmds)

ecramid_site_scores <- 
  data.frame(scores(ecramid.nmds)$sites) |> 
  mutate(Site = rownames(scores(ecramid.nmds)$sites))  
  #separate(Site, into = c("invasive", "Final_ID"), sep = "_")


ecramid_species_scores <- 
  data.frame(scores(ecramid.nmds)$species) |>
  mutate(Species = rownames(scores(ecramid.nmds)$species))
```

Plots

```{r}
ecramid_NMDS_plot <- ggplot() + 
  geom_text(data = ecramid_site_scores, aes(x = NMDS1, y = NMDS2, label = Site, color = Site), cex = 3) + 
  geom_point(data = ecramid_species_scores, aes(x = NMDS1, y = NMDS2, label = Species))

ecramid_NMDS_plot
```

plotting with hulls

```{r}
grp.Nat <- ecramid_site_scores[ecramid_site_scores$Native == "NATIVE", ][chull(ecramid_site_scores[ecramid_site_scores$Native == 
    "NATIVE", c("NMDS1", "NMDS2")]), ]  # hull values for grp A
grp.NonNat <- ecramid_site_scores[Site_scores$Native == "NN", ][chull(Site_scores[Site_scores$Native == 
    "NN", c("NMDS1", "NMDS2")]), ]  # hull values for grp B

hull.data <- rbind(grp.Nat, grp.NonNat)  #combine grp.a and grp.b
hull.data


NMDS_plot + 
  geom_polygon(data = hull.data, aes(x = NMDS1, y = NMDS2, color = Native, fill = Native), alpha = 0.1)
```

#potential later maps - color-code site locations by pre-fire/post-fire to compare

DID SOMEONE ON THE TEAM ADD SOME FILES?

1.  Switch to main and "Pull"
2.  Switch BACK to IZZYs branch and run:

git fetch -–all

git fetch origin/main
