---
title: "Veg_Analysis"
format: html
date created: 2/22/2024
editor: visual
---

# Vegetation Analysis

Izzy Funes

### Load the Primary Dataframe

```{r}
load("CRAM_Plant_Primary.Rdata")
CRAM_envData <- read.csv("CRAM_envData.csv")
TaxizedPlants <- read.csv("TaxizedPlants3-11.csv")
```

### Load packages

```{r}
library(tidyverse)
library(readxl)
library(vegan)
#install.packages("ggnewscale")
library(ggnewscale)
library(factoextra)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(lubridate)
library(reshape2)
library(ggplot2)
library(ggpubr)
library(ggvegan)
```

## Create Matrix

```{r}
Veg_Primary <- CRAM_PlantJoin 
#might not need this though

Plant_and_Fire_Join <- inner_join(CRAM_envData, TaxizedPlants)
#cleaner data join
Plant_and_Fire_Species <- Plant_and_Fire_Join |>
  select(Final_ID, ecramid, PrePost)
#Same as CRAM_sites_species
Invasive_Site <- Plant_and_Fire_Join |>
  select(ecramid, invasive, PrePost)
#Same as SpeciesEnv
Species_Inv <- Plant_and_Fire_Join |>
  select(invasive, Final_ID, PrePost) |>
  distinct()

test <- table(Plant_and_Fire_Species[1:2])

test <- as.data.frame(test)


PF_Freq_matrix <- test |> 
  mutate(Present = 1) |> 
  pivot_wider(
    id_cols = ecramid,
    names_from = Final_ID, 
    values_from = Freq, 
    values_fn = sum, 
    values_fill = 0)
     #takes columns and turns them into rows which shortens the number of rows (makes into dataframe format)
PF_Freq_matrix

PF_Remove_ecramid_matrix <- PF_Freq_matrix |>
  select(-ecramid)

PF_Remove_ecramid_matrix

#Species_matrix <- PF_wide[2:ncol(PF_Freq)] |> 
  #as.matrix()

#rownames(Species_matrix) <- PF_Freq$ecramid
#Species_matrix

#PF_Presence <- Plant_and_Fire_Clean |> 
  #mutate(Present = 1) |> 
  #pivot_wider(names_from = Final_ID, values_from = Present, values_fn = sum, values_fill = 0)

#Species_Transpose <- t(Species_List) 
#View(Species_matrix)
```

### Column count (for funsies?)

```{r}
 Plant_and_Fire_Species|>
  group_by(ecramid) |>
  count()
```

## Community Diversity Analysis

```{r}
#Try Jaccard or Bray-Curtis index

transpose <- t(PF_Remove_ecramid_matrix)

dist(transpose, method = "euclidean")

vegdist(transpose, method = "bray")
```

Kind of long output.....not sure how to interpret

## NMDS Analysis Pt. 1

### Garrett's Way

```{r}
#ecramid.nmds <- metaMDS(Species_matrix, distance = "man")
# same options for distance as vegdist
Veg.PA <- decostand(PF_Remove_ecramid_matrix, method = "pa")
#veg.PA same as veg.hel
veg.nmds <- metaMDS(Veg.PA, autotransform = F,trymax=100)

ordiplot(veg.nmds, type = "t") 

plot(veg.nmds, type = "p", display = c("site", "species"))


#remotes::install_github("gavinsimpson/ggvegan", force = TRUE)

autoplot(veg.nmds)
#veg.nmds.df <- data.frame(veg.nmds)
#can't turn into a data frame apparently 
fort <- fortify(veg.nmds)


scores <- scores(veg.nmds)

scores

```

### ggplots

```{r}
p1 <- ggplot()+ 
  geom_point(data = subset(fort, score == "sites"),
             mapping = aes(x = NMDS1, y = NMDS2),
             color = "black",
             alpha = 0.5)+
  geom_segment(data = subset(fort, score == "species"),
                             mapping = aes (x = 0, y = 0, xend = NMDS1, yend = NMDS2),
                             arrow = arrow(length = unit(0.015, "npc"),
                                           type = "closed"),
                             color = "darkgray",
                             size = 0.8)+
  geom_text(data = subset(fort, score == 'species'),
            mapping = aes(label = label, x = NMDS1 *1.1, y = NMDS2 *1.1))+
  geom_abline(intercept = 0, slope = 0, linetype="dashed", size = 0.8, color = "gray")+
  geom_vline(aes(xintercept = 0), linetype = "dashed",  size = 0.8, color = "gray")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = "black"))
p1
```

```{r}
p2 <- ggplot()+ 
  geom_point(data = subset(fort, score == "sites"),
             mapping = aes(x = NMDS1, y = NMDS2),
             color = "black",
             alpha = 0)+
  geom_segment(data = subset(fort, score == "species"),
               mapping = aes (x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               arrow = arrow(length = unit(0.015, "npc"),
                             type = "closed"),
               color = "darkgray",
               size = 0.8)+
  geom_text(data = subset(fort, score == 'species'),
            mapping = aes(label = label, x = NMDS1 *1.1, y = NMDS2 *1.1))+
  geom_abline(intercept = 0, slope = 0, linetype="dashed", size = 0.8, color = "gray")+
  geom_vline(aes(xintercept = 0), linetype = "dashed",  size = 0.8, color = "gray")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = "black"))

p2


ggarrange(p1, p2, ncol = 1)


```

```{r}
#summary(CRAM_envData)
p3 <-  ggplot()+ 
  geom_point(data = subset(fort, score == "sites"),
             mapping = aes(x = NMDS1, y = NMDS2, color = CRAM_envData$PrePost ),
             alpha = 0.5)+
  geom_segment(data = subset(fort, score == "species"),
               mapping = aes (x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               arrow = arrow(length = unit(0.015, "npc"),
                             type = "closed"),
               color = "darkgray",
               size = 0,
               alpha = 0)+
  geom_text(data = subset(fort, score == 'species'),
            mapping = aes(label = label, x = NMDS1 *1.1, y = NMDS2 *1.1),
            alpha = 0)+
  geom_abline(intercept = 0, slope = 0, linetype="dashed", size = 0.8, color = "gray")+
  geom_vline(aes(xintercept = 0), linetype = "dashed",  size = 0.8, color = "gray")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = "black"))+
  guides(color = guide_legend(title = "Pre or Post Fire")) 
p3


p4 <- ggplot()+ 
  geom_point(data = subset(fort, score == "sites"),
             mapping = aes(x = NMDS1, y = NMDS2),
             color = "black",
             alpha = 0)+
  geom_segment(data = subset(fort, score == "species"),
               mapping = aes (x = 0, y = 0, xend = NMDS1, yend = NMDS2, color = Invasive_Site$invasive),
               arrow = arrow(length = unit(0.015, "npc"),
                             type = "closed"),
               size = 0.8)+
  geom_text(data = subset(fort, score == 'species'),
            mapping = aes(label = label, x = NMDS1 *1.1, y = NMDS2 *1.1, color = Invasive_Site$invasive))+
  geom_abline(intercept = 0, slope = 0, linetype="dashed", size = 0.8, color = "gray")+
  geom_vline(aes(xintercept = 0), linetype = "dashed",  size = 0.8, color = "gray")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = "black"))
p4



ggarrange(p3, p4, ncol = 1)
```

```{r}
ggplot()+
  geom_point(data = CRAM_envData, aes(x = DaysSinceFire, y = FireDist, color = PrePost))+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = "black"))
```

Statistical Analysis

```{r}
x <- adonis2(Veg.PA~DaysSinceFire*FireDist, data = CRAM_envData)

z <- adonis2(Veg.PA~DaysSinceFire*FireDist*Elevation*FireSize, data = CRAM_envData)

y <- adonis2(Veg.PA~FireDist, data = CRAM_envData)
```

## NMDS Analysis PT. 2

### Izzy's Way

```{r}
veg.PA.1 <- decostand(PF_Remove_ecramid_matrix, method = "pa")
veg.nmds.1  <- metaMDS(Veg.PA.1, distance = "man")
```

Extract scores

Insert fire history data

native/non-native is "invasive" score (0/1)

filter in site data (comparing pre-fire and post-fire)

```{r}
scores(veg.nmds.1)

ecramid_site_scores <- 
  data.frame(scores(veg.nmds.1)$sites)|>
  mutate(Site = rownames(scores(veg.nmds)$sites))  
  #separate(Site, into = c("invasive", "Final_ID"), sep = "_")

sites_join <- inner_join(ecramid_site_scores, Invasive_Site)

ecramid_species_scores <- 
  data.frame(scores(veg.nmds.1)$species) |>
  mutate(Species = rownames(scores(veg.nmds.1)$species))
```

### ggplots

```{r}
ecramid_NMDS_plot <- ggplot() + 
  geom_text(data = subset(fort, score == 'species'),
            mapping = aes(label = label, x = NMDS1 *1.1, y = NMDS2 *1.1),
            alpha = 0) + 
  geom_point(data = subset(fort, score == "sites"),
             mapping = aes(x = NMDS1, y = NMDS2),
             color = "black",
             alpha = 0)

ecramid_NMDS_plot
```

```{r}
NMDS_plot <- ggplot() + 
  geom_text(data = fort, aes(x = NMDS1, y = NMDS2, label = Species, color = Native), cex = 3) + 
  geom_point(data = Species_scores, aes(x = NMDS1, y = NMDS2, label = Species))
NMDS_plot
```

plotting with hulls

```{r}
grp.Nat <- ecramid_site_scores[ecramid_site_scores$Native == "NATIVE", ][chull(ecramid_site_scores[ecramid_site_scores$Native == 
    "NATIVE", c("NMDS1", "NMDS2")]), ]  # hull values for grp A
grp.NonNat <- ecramid_site_scores[Site_scores$Native == "NN", ][chull(Site_scores[Site_scores$Native == 
    "NN", c("NMDS1", "NMDS2")]), ]  # hull values for grp B

hull.data <- rbind(grp.Nat, grp.NonNat)  #combine grp.a and grp.b
hull.data


NMDS_plot + 
  geom_polygon(data = hull.data, aes(x = NMDS1, y = NMDS2, color = Native, fill = Native), alpha = 0.1)
```

#potential later maps - color-code site locations by pre-fire/post-fire to compare

DID SOMEONE ON THE TEAM ADD SOME FILES?

1.  Switch to main and "Pull"
2.  Switch BACK to IZZYs branch and run:

git fetch -–all

git fetch origin/main
